---
title: "Toothfish analysis simulation testing"
author: "Stephen D. Gregory"
format: 
  html:
    self-contained: true
    code-fold: show
    toc: true
lightbox: true 
---

## A simulation function

```{r}
#| label: sim-fun

# simulate expected lengths function
sim_exp_len_f <- function(data, 
                          lp_formula, lp_coefs, lp_parms = NULL) {
  
  # get formula
  f_ <- as.formula(object = lp_formula)
  
  # get lp coefficients
  c_ <- sapply(X = names(lp_coefs),
               FUN = function(v) {
                 grepl(pattern = v,
                       x = gsub(pattern = "^1 \\+",
                                replacement = "intercept +",
                                x = as.character(f_)[[2]]))
               })
  
  # get data
  d_ <- data.frame(data)[, names(which(c_))]
  
  # add delta to data
  if (grepl("delta", as.character(f_)[[2]])) {
    if (grepl("delta\\[", as.character(f_)[[2]])) {
      d_$delta <- lp_parms$delta[data$sex]
    } else {
      d_$delta <- lp_parms$delta[1]
    }
  }
  
  # add gamma to data
  if(grepl("gamma", as.character(f_)[[2]])) {
    d_$gamma <- lp_parms$gamma
  }
  
  # make the model matrix
  mm <- model.matrix(object = f_, data = d_)
  
  # get the expectations
  mm %*% lp_coefs[c_]
  
}

```

## Some examples

### Example 1: intercept, sex & age

```{r}
#| label: sim-fun-eg1

# example 1: intercept, age & sex
lp_f <- ~ 1 + sex + age
lp_c <- c("intercept" = 50,
          "sex" = 20,
          "age" = 2)
d <- data.frame("intercept" = 1, 
                "sex" = factor(c(1, 2)),
                "age" = c(20, 20))
sim_exp_len_f(data = d, lp_formula = lp_f, lp_coefs = lp_c)

```

### Example 2: dm with breakpoint only

```{r}
#| label: sim-fun-eg2

# example 2: dm with breakpoint only
lp_f <- ~ 0 + I(dm1 - delta) +  I(sqrt((dm2 - delta)^2 + gamma))
lp_c <- c("dm1" = 2.5, 
          "dm2" = -1.7)
lp_p <- list("delta" = 15,
             "gamma" = 0.3)
d <- data.frame("dm1" = c(50, 50), 
                "dm2" = c(25, 25))
sim_exp_len_f(data = d, lp_formula = lp_f, lp_coefs = lp_c, lp_parms = lp_p)

```

### Example 3: sex, age & dm with breakpoint

```{r}
#| label: sim-fun-eg3

# example 3: intercept, age, sex & dm with breakpoint
lp_f <- ~ 1 + sex + age + I(dm1 - delta) + I(sqrt((dm2 - delta)^2 + gamma))
lp_c <- c("intercept" = 50,
          "sex" = 20,
          "age" = 2,
          "dm1" = 2.5, 
          "dm2" = -1.7)
lp_p <- list("delta" = 15,
             "gamma" = 0.3)
d <- data.frame("intercept" = 1,
                "sex" = factor(c(1, 2)),
                "age" = c(20, 20), 
                "dm1" = c(50, 50),
                "dm2" = c(25, 25))
sim_exp_len_f(data = d, lp_formula = lp_f, lp_coefs = lp_c, lp_parms = lp_p)

```

### Example 4: intercept, age, sex & dm with sex-specific breakpoints

```{r}
#| label: sim-fun-eg4

# example 4: intercept, age, sex & dm with sex-specific breakpoints
lp_f <- ~ 1 + sex + age + I(dm1 - delta[sex]) + I(sqrt((dm2 - delta[sex])^2 + gamma))
lp_c <- c("intercept" = 50,
          "sex" = 20,
          "age" = 2,
          "dm1" = 2.5, 
          "dm2" = -1.7)
lp_p <- list("delta" = c(10, 20),
             "gamma" = 0.3)
d <- data.frame("intercept" = 1, 
                "sex" = factor(c(1, 2)),
                "age" = c(20, 20),
                "dm1" = c(50, 50),
                "dm2" = c(25, 25))
sim_exp_len_f(data = d, lp_formula = lp_f, lp_coefs = lp_c, lp_parms = lp_p)

```

### Example 5: intercept, age, sex & dm

```{r}
#| label: sim-fun-eg5

# example 5: intercept, age, sex & dm
lp_f <- ~ 1 + sex + age + dm
lp_c <- c("intercept" = 50,
          "age" = 2,
          "sex" = 20,
          "dm" = 2.5)
d <- data.frame("intercept" = 1,
                "sex" = factor(c(1, 2)),
                "age" = c(20, 20),
                "dm" = c(50, 50))
sim_exp_len_f(data = d, lp_formula = lp_f, lp_coefs = lp_c)

```

## Simulation set-up

```{r}
#| label: sim-setup
#| message: false
#| warning: false

# libraries
library(data.table)
library(dclone)
library(ggplot2)
library(patchwork)
library(gt)
library(bayesplot)

# MCMC numbers
mc <- list(n.chains = 3, 
           n.adapt = 1000, 
           n.update = 2000, 
           thin = 10, 
           n.iter = 13000)

```

## Simulate explanatory variables

```{r}
#| label: simulate-explanatories

# set seed
set.seed(4)

# constants
n <- 3600
z <- 1.5:30
earliest_birth <- as.Date("1993-07-01")
latest_capture <- as.Date("2024-02-21")
yr_diff <- year(latest_capture) - year(earliest_birth)

# temperature data, allowing for possible trend
a <- 0.5
p <- c("months" = 12) # c("days" = 365)
temp <- a * cos(x = 2 * pi * (1:(p * yr_diff)) / p) + 
  a * sin(x = 2 * pi * (1:(p * yr_diff)) / p) + 
  rnorm(n = p * yr_diff,
        mean = seq(from = 1.02, to = 1.02, length.out = 365 * yr_diff),
        sd = 0.5)
temp_dt <- data.table(
  "date" = seq.Date(from = as.Date(paste0(year(earliest_birth), "-03-01")), 
                    to = as.Date(paste0(year(latest_capture), "-02-28")), 
                    by = names(p))[1:length(temp)],
  "temperature" = temp)
temp_dt[, "month" := month(date)]
temp_dt[, "year" := year(date)]
temp_p <- ggplot(data = temp_dt, mapping = aes(x = date, y = temperature)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = as.Date(x = paste0(year(earliest_birth):
                                               (year(latest_capture) - 1),
                                             "-03-01")),
             linewidth = 0.5, colour = "grey") +
  theme_classic()

# explanatory variables: age & sex
age <- sample(x = z, size = n, replace = TRUE, 
              prob = plogis(0.02 * z)) * 365
sex <- gl(n = 2, k = 1, length = n)

# explanatory variables: degree months
lh_dates <- data.table(
  "id" = 1:n,
  "age" = age,
  "capture" = as.Date(sapply(X = 1:n,
                             FUN = function(v) {
                               sample(x = seq.Date(from = (earliest_birth + age[v]),
                                                   to = latest_capture, 
                                                   by = "days"), size = 1)}))
)
lh_dates[, "birth" := capture - age]
lh_temps <- temp_dt[lh_dates, on = .(date > birth, date < capture)]
dm <- lh_temps[which(temperature > 0), mean(temperature), 
               by = .(id, year, month)][, sum(V1), by = id]$V1

# pull them together; note name of age & dm data
d <- data.table(intercept = 1, 
                "sex" = factor(sex), 
                "age" = age,  
                "dm" = dm)

# convert age to months
d[, "age" := floor((age / 365) * 12)]

# add replicate age column - needed for simulation
d[, c("age1", "age2") := age]
d[, "age" := NULL]

# remove fish age > 30 years
d <- subset(x = d, subset = age1 < (30 * 365) / 12)

# show that age & dm are not independent
temp_dm <- ggplot(data = d, mapping = aes(x = age1, y = dm)) +
  geom_point() +
  ggtitle(label = paste0("Pearson's correlation = ", 
                         sprintf(with(data = d, expr = cor(age1, dm)), 
                                 fmt = "%.3f"))) +
  theme_classic()

# make dm an anomaly
d[, "dma" := (dm - age1)]

# add replicate dma column - needed for simulation
d[, c("dma1", "dma2") := dma]
d[, "dma" := NULL]

# show that age & dm are more independent
temp_dma <- ggplot(data = d, mapping = aes(x = age1, y = dma1)) +
  geom_point() +
  ggtitle(label = paste0("Pearson's correlation = ", 
                         sprintf(with(data = d, expr = cor(age1, dma1)), 
                                 fmt = "%.3f"))) +
  theme_classic()

# combine plots
temp_dm_p <- temp_p / (temp_dm + temp_dma)

# plot
print(temp_dm_p)

```


## Model M0

```r 
{{< include ../../models/M0.jags >}}
```

### Simulate response variable

```{r}
#| label: m0-simulate-response

# linear predictor (lp) coefficients
m0_coefs <- c("intercept" = 50,
              "sex" = -10.0,
              "age1" = 0.5)

# model error
sd <- 0.1

# simulate expected lengths
lp <- sim_exp_len_f(data = d,
                    lp_formula = ~ 1 + sex + age1, 
                    lp_coefs = m0_coefs)

# simulate actual lengths
lengths <- rnorm(n = nrow(d), mean = lp, sd = sd * d$age1)

```

### Plot

```{r}
#| label: m0-repsonse-plot

# plot response variable
m0_sim_p <- ggplot(data = cbind(d, lengths), 
                   mapping = aes(y = lengths, x = age1, colour = `sex`)) +
  geom_point() +
  facet_wrap(~ sex) + 
  labs(x = "Age (months)", y = "Length (cm)") +
  theme(text = element_text(size = 16), strip.text = element_text(size = 10),
        legend.position = "none")
print(m0_sim_p)

```

### Fit

```{r}
#| label: m0-fit
#| message: false
#| warning: false
#| results: hide

# fit
m0_fit <- jags.fit(data = list("n" = nrow(d), 
                               "length" = lengths, 
                               "sex" = d[, sex], 
                               "age" = d[, age1]),
                   params = c("intercept", "beta_sex", "beta_age", "sigma", "tau"),
                   model = "../../models/M0.jags",
                   n.chains = mc$n.chains, 
                   n.adapt = mc$n.adapt, 
                   n.update = mc$n.update, 
                   thin = mc$thin, 
                   n.iter = mc$n.iter)

# get the estimates
m0_ests <- with(summary(m0_fit), 
                as.data.table(cbind(statistics[, 1, drop = FALSE], 
                                    quantiles[, c("2.5%", "97.5%")]), 
                              keep.rownames = "Coefficient")
)

# add actual
idx <- match(x = gsub(pattern = "1$|2$", replacement = "", 
                      x = names(m0_coefs)),
             table = gsub(pattern = "beta_|\\[2\\]", replacement = "",
                          x = m0_ests$Coefficient))
m0_ests[idx, "Actual"] <- m0_coefs

```


## Model M1

```r 
{{< include ../../models/M1.jags >}}
```

### Simulate response variable

```{r}
#| label: m1-simulate-response

# linear predictor (lp) coefficients
m1_coefs <- c("intercept" = 100,
              "sex" = -10.0,
              "age1" = 0.5,
              "age2" = -0.2)

# lp parameters
m1_parms <- list("delta" = c(115, 100),
                 "gamma" = 25)

# model error
sd <- 0.1

# simulate expected lengths
lp <- sim_exp_len_f(data = d,
                    lp_formula = ~ 1 + sex + I(age1 - delta[sex]) + I(sqrt((age2 - delta[sex])^2 + gamma)), 
                    lp_coefs = m1_coefs,
                    lp_parms = m1_parms)

# simulate actual lengths
lengths <- rnorm(n = nrow(d), mean = lp, sd = sd * d$age1)

```

### Plot

```{r}
#| label: m1-repsonse-plot

# plot response variable
m1_sim_p <- ggplot(data = cbind(d, lengths), 
                   mapping = aes(y = lengths, x = age1, colour = `sex`)) +
  geom_point() +
  facet_wrap(~ sex) + 
  labs(x = "Age (months)", y = "Length (cm)") +
  theme(text = element_text(size = 16), strip.text = element_text(size = 10),
        legend.position = "none")
print(m1_sim_p)

```

### Fit

```{r}
#| label: m1-fit
#| message: false
#| warning: false
#| results: hide

# fit
m1_fit <- jags.fit(data = list("n" = nrow(d), 
                               "length" = lengths, 
                               "sex" = d[, sex], 
                               "age" = d[, age1],
                               "min_age" = min(d$age1),
                               "max_age" = max(d$age1)),
                   params = c("intercept", "beta_sex", 
                              "beta_age1", "beta_age2", 
                              "delta", "gamma", 
                              "sigma", "tau"),
                   model = "../../models/M1.jags",
                   n.chains = mc$n.chains, 
                   n.adapt = mc$n.adapt, 
                   n.update = mc$n.update, 
                   thin = mc$thin, 
                   n.iter = mc$n.iter)

# get the estimates
m1_ests <- with(summary(m1_fit), 
                as.data.table(cbind(statistics[, 1, drop = FALSE], 
                                    quantiles[, c("2.5%", "97.5%")]), 
                              keep.rownames = "Coefficient")
)

# add actual
idx <- match(x = names(m1_coefs),
             table = gsub(pattern = "beta_|\\[2\\]", replacement = "",
                          x = m1_ests$Coefficient))
m1_ests[idx, "Actual"] <- m1_coefs
idx <- match(x = names(unlist(m1_parms)),
             table = gsub(pattern = "\\[([0-9])\\]", replacement = "\\1",
                          x = m1_ests$Coefficient))
m1_ests[idx, "Actual"] <- unlist(m1_parms)

```


## Model M2

```r 
{{< include ../../models/M2.jags >}}
```

### Simulate response variable

```{r}
#| label: m2-simulate-response

# linear predictor (lp) coefficients
m2_coefs <- c("intercept" = 50,
              "sex" = -10.0,
              "age1" = 0.5,
              "dma1" = -0.2)

# model error
sd <- 0.1

# simulate expected lengths
lp <- sim_exp_len_f(data = d,
                    lp_formula = ~ 1 + sex + age1 + dma1, 
                    lp_coefs = m2_coefs)

# simulate actual lengths
lengths <- rnorm(n = nrow(d), mean = lp, sd = sd * d$age1)

```

### Plot

```{r}
#| label: m2-repsonse-plot

# plot response variable
m2_sim_p <- ggplot(data = cbind(d, lengths), 
                   mapping = aes(y = lengths, x = age1, colour = `sex`)) +
  geom_point() +
  facet_wrap(~ sex) + 
  labs(x = "Age (months)", y = "Length (cm)") +
  theme(text = element_text(size = 16), strip.text = element_text(size = 10),
        legend.position = "none")
print(m2_sim_p)

```

### Fit

```{r}
#| label: m2-fit
#| message: false
#| warning: false
#| results: hide

# fit
m2_fit <- jags.fit(data = list("n" = nrow(d), 
                               "length" = lengths, 
                               "sex" = d[, sex], 
                               "age" = d[, age1],
                               "dm" = d[, dma1]),
                   params = c("intercept", "beta_sex", 
                              "beta_age", "beta_dm", 
                              "sigma", "tau"),
                   model = "../../models/M2.jags",
                   n.chains = mc$n.chains, 
                   n.adapt = mc$n.adapt, 
                   n.update = mc$n.update, 
                   thin = mc$thin, 
                   n.iter = mc$n.iter)

# get the estimates
m2_ests <- with(summary(m2_fit), 
                as.data.table(cbind(statistics[, 1, drop = FALSE], 
                                    quantiles[, c("2.5%", "97.5%")]), 
                              keep.rownames = "Coefficient")
)

# add actual
idx <- match(x = gsub(pattern = "1$|2$", replacement = "", 
                      x = names(m2_coefs)),
             table = gsub(pattern = "beta_|\\[2\\]", replacement = "",
                          x = m2_ests$Coefficient))
m2_ests[idx, "Actual"] <- m2_coefs
idx <- match(x = "beta_dm", table = m2_ests$Coefficient)
m2_ests[idx, "Actual"] <- m2_coefs["dma1"]

```


## Model M3

```r 
{{< include ../../models/M3.jags >}}
```

### Simulate response variable

```{r}
#| label: m3-simulate-response

# linear predictor (lp) coefficients
m3_coefs <- c("intercept" = 100,
              "sex" = -10.0,
              "age1" = 0.5,
              "age2" = -0.2,
              "dma1" = -0.2)

# lp parameters
m3_parms <- list("delta" = c(115, 100),
                 "gamma" = 25)

# model error
sd <- 0.1

# simulate expected lengths
lp <- sim_exp_len_f(data = d,
                    lp_formula = ~ 1 + sex + I(age1 - delta[sex]) + I(sqrt((age2 - delta[sex])^2 + gamma)) + dma1, 
                    lp_coefs = m3_coefs,
                    lp_parms = m3_parms)

# simulate actual lengths
lengths <- rnorm(n = nrow(d), mean = lp, sd = sd * d$age1)

```

### Plot

```{r}
#| label: m3-repsonse-plot

# plot response variable
m3_sim_p <- ggplot(data = cbind(d, lengths), 
                   mapping = aes(y = lengths, x = age1, colour = `sex`)) +
  geom_point() +
  facet_wrap(~ sex) + 
  labs(x = "Age (months)", y = "Length (cm)") +
  theme(text = element_text(size = 16), strip.text = element_text(size = 10),
        legend.position = "none")
print(m3_sim_p)

```

### Fit

```{r}
#| label: m3-fit
#| message: false
#| warning: false
#| results: hide

# fit
m3_fit <- jags.fit(data = list("n" = nrow(d), 
                               "length" = lengths, 
                               "sex" = d[, sex], 
                               "age" = d[, age1],
                               "dm" = d[, dma1],
                               "min_age" = min(d$age1),
                               "max_age" = max(d$age1)),
                   params = c("intercept", "beta_sex", 
                              "beta_age1", "beta_age2",
                              "beta_dm", 
                              "delta", "gamma", 
                              "sigma", "tau"),
                   model = "../../models/M3.jags",
                   n.chains = mc$n.chains, 
                   n.adapt = mc$n.adapt, 
                   n.update = mc$n.update, 
                   thin = mc$thin, 
                   n.iter = mc$n.iter)

# get the estimates
m3_ests <- with(summary(m3_fit), 
                as.data.table(cbind(statistics[, 1, drop = FALSE], 
                                    quantiles[, c("2.5%", "97.5%")]), 
                              keep.rownames = "Coefficient")
)

# add actual
idx <- match(x = names(m3_coefs),
             table = gsub(pattern = "beta_|\\[2\\]", replacement = "",
                          x = m3_ests$Coefficient))
m3_ests[idx, "Actual"] <- m3_coefs
idx <- match(x = names(unlist(m3_parms)),
             table = gsub(pattern = "\\[([0-9])\\]", replacement = "\\1",
                          x = m3_ests$Coefficient))
m3_ests[idx, "Actual"] <- unlist(m3_parms)
idx <- match(x = "beta_dm", table = m3_ests$Coefficient)
m3_ests[idx, "Actual"] <- m3_coefs["dma1"]

```


## Table of coefficients

```{r}
#| label: coef-tab
#| results: asis

# get model estimates
ests_lst <- lapply(X = ls(pattern = "_ests"), 
                   FUN = function(v) eval(as.name(v)))
names(ests_lst) <- tools::toTitleCase(gsub(pattern = "_ests", 
                                           replacement = "", 
                                           x = ls(pattern = "_ests")))

# make table
ests_dt <- rbindlist(l = ests_lst, idcol = "Model")

# make a gt
ests_gt <- gt(data = ests_dt)

# blank repeat model names to be cleaner
ests_gt$`_data`[duplicated(ests_gt$`_data`$Model), "Model"] <- ""

# make font size
ests_gt <- tab_options(data = ests_gt, table.font.size = 12)

# make first row bold
lst <- lapply(colnames(ests_dt), function(v) md(paste0("**", v, "**")))
names(lst) <- colnames(ests_dt)
ests_gt <- cols_label(.data = ests_gt,
                      .list = lst)

# change numeric digits
ests_gt <- fmt_number(data = ests_gt, decimals = 3)

# print
ests_gt

```


## Coefficient plots

```{r}
#| label: coef-plot
#| fig-width: 8
#| fig-height: 5
#| message: false
#| warning: false

# make plotting nicer
ests_dt[, "Type" := "betas"]
ests_dt[grep(pattern = "delta\\[|delta_sex\\[", x = Coefficient), 
        "Type" := "deltas"]

# plot
coef_p <- ggplot(data = subset(x = ests_dt, 
                               subset = !Coefficient %in% c("sigma", "tau")),
                 mapping = aes(x = Coefficient, y = Mean,
                               ymin = `2.5%`, ymax = `97.5%`)) +
  geom_point() +
  geom_errorbar(width = 0.25) +
  geom_point(data = ests_dt, mapping = aes(y = Actual), colour = "red") +
  facet_grid(cols = vars(Model), rows = vars(Type), scales = "free") +
  labs(x = "Model parameter", y = "Estimate / generating value") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
print(coef_p)

```


## Example diagnostics

```{r}
#| label: eg-traces
#| fig-width: 8
#| fig-height: 5
#| message: false
#| warning: false

# trace plot
trace_p <- mcmc_trace(m3_fit)
print(trace_p)

```

```{r}
#| label: eg-pairs
#| fig-width: 12
#| fig-height: 12
#| message: false
#| warning: false

# pairs plot
pairs_p <- mcmc_pairs(m3_fit)
print(pairs_p)

```

```{r}
#| include: false

# save
save.image("toothfish-analysis-simulations.RData")

```

```{r}
#| include: false

# save figure 2


```
